Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-04-30 src/tex/ms.tex
[showyourwork] overleaf sync

    \subsection{Related Work}
    In literature there have been multiple efforts to correct the degrading of the images using statistical methods.
    The standard baseline method for image deconvolution was proposed independently by \cite{Richardson1972} and \cite{Lucy1974}. They used expectation maximisation to solve. We will refer to this method as "RL".

    One of the significant challenges in image deconvolution is the problem of Rayleigh-Lynden Bell (RL) decomposition of structures into point sources. This problem was first introduced by Reeves et al. \cite{Reeves1994} and later analyzed by Fish et al. \cite{Fish1995}. Regularized RL methods, such as total variation \citep{Dey2006} and maximum entropy \citep{Bi1994}, have been proposed to overcome this issue.

    Multi-scale LIRA prior has been used in image deconvolution to incorporate error estimates and Bayesian sampling. This prior has been used in various studies, including those by Esch et al. \cite{Esch2004} and Connors et al. \cite{Connors2011}. However, the prior is ad-hoc and lacks physical information or assumptions, limiting its applicability.

@@

    An alternative approach to solve the problem in the framework of information field theory.

    Recently, image deconvolution using physical assumptions and decomposition into point and diffuse flux in 3D/4D has gained attention. Pumpe et al. \cite{Pumpe2018} and Selig et al. \cite{Selig2015} proposed priors based on physical assumptions, which have been shown to work well for large-scale emission. However, the smoothness assumption in these methods can lead to artifacts in the galactic plane.

@@

    \begin{itemize}
         \item  Problem of RL decomposing structures into point sources: \cite{Reeves1994} \cite{Fish1995}
         \item regularised RL , e.g. total variation \citep{Dey2006} or maximum entropy \citep{Bi1994} \todo{other references?}
        \item Joint likelihood RL: \cite{Ingaramo2014} has found that RL take the best out of each image and create a "merged" image
        \item Multi-scale LIRA prior of course: \cite{Esch2004, Connors2011}, advantage of error estimates / Bayesian sampling. However prior is rather "ad-hoc", no physical information / assumptions. Just "smoothness"
        \item What abput d3po and d4po? Priors based on physical assumptions, decomposition into point and diffuse flux, but in 3d/4d \cite{Selig2015}\cite{Pumpe2018}. The smoothness assumption works for large scale emission, but leads to artifacts in the galactic plane... 
    \end{itemize}

@@

    \subsection{Some notes on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the \textit{de facto} standard for de-blurring and super-resolution of natural images \todo{reference review?}. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

@@

    \todo{this sum corresponds to the loop over data sets in machine learning... }

@@

    \subsection{Patch Based Priors}
    In general it is difficult to capture global image statistics in a prior likelihood
    taking into account all pixel to pixel correlations. The size of the corresponding
    correlation matrix would grow with the number of pixels squared and thus quickly
    becomes computationally untraceable. However astronomical images typically only
    show local correlations between pixels. This is because the spatial
    scale of astrophysical processes in an image is always limited by the finite
    interaction time and the fact that the interaction cannot propagate
    faster than the speed of light (structure in the CMB is an exception here). This means on smaller scales astronomical
    images often contain basic structures such as edges, corners, filaments or
    periodic patterns formed by local astrophysical processes.

    \todo{This is interesting, but move to intro...}
    \cite{Zoran2011} build on very similar assumption for natural images.
    Instead of assuming a fixed local correlation structures, such as smoothness
    they proposed to learn the local correlation structure on millions of patches
    extracted from natural images.

    One of the main advantages of the GMM is the possibility to evaluate its likelihood
    in closed form. For a given single patch $\mathbf{x}_n$ it is given by:

    \begin{equation}
        \label{eq:gmm}
        P_{GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k N(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2)
    \end{equation}

    Where $N$ is a Gaussian distribution, $\boldsymbol{\mu}_k$ the vector of means for
    a given component $k$, $\boldsymbol{\Sigma}_k$ the corresponding covariance matrix
    and $\pi_k$ the weight the component. The different weights are required to sum to unity such that $\sum_{k=1}^N\pi_k = 1$.

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch
    from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the GMM component becomes a hyper-parameter in the model and is optimized along with the image $\mathbf{x}$.

@@

    \cite{Bouman2016} later adapted the patch prior reconstruction to be used
    with radio astronomy data.
    They have shown that the reconstructed image only weakly depends on the choice
    of the reference data on which the the patch prior is learned. They found
    equivalent results for GMMs learned on natural images and specifically
    simulated images of black hole ring structures.

@@

    We also use \texttt{Numpy}~\citep{Numpy2020} for handling data arrays and
    \texttt{Matplotlib}~\citep{Hunter2007} for plotting.
    To learn the Gaussian Mixture Models from patches we the standard implementation
    provided in the \texttt{Scikit-Learn}~\citep{Skimage2014} package. To handle the 
    serialisation of images to the FITS data format and to handle world coordinate
    systems (WCS) we use \texttt{Astropy}~\citep{Astropy2018}.
    For optimization and internal data handling we use \texttt{PyTorch}.

@@

    \vlk{We have devised four distinct arrangements of combinations of point and extended sources of differents shapes to test the algorithms.  The arrangements are 
    \begin{itemize}
        \item[(A)] {\bf asterism:} A set of four point sources arranged around an extended Gaussian source
        \item[(B)] {\bf points:} Point sources arranged with different separations
        \item[(C)] {\bf shield:} Disk shaped flat extended source with superposed point sources and linear jet-like features
        \item[(D)] {\bf spiral:} Extended thin double-spiral with a flat disk at the center and point sources adjacent to it
    \end{itemize}
    The arrangements are shown in Figure~TBD, and the details of the locations and brightness of each component is listed in Table~TBD.}

@@

    \subsection{Methods}
    \label{subsec:methods}
    We compare the performance of \jolideco with different prior assumptions and against the \textit{LIRA} method introduced by \cite{Esch2004}. More specifically we use the following definitions:

@@

        \item \textbf{Jolideco (Patch, GLEAM v0.1)}: \jolideco with a patch prior learned on GLEAM data, described in Section~\ref{sssec:gleam-radio-data}. It also uses the standardised patch normalisation as described by Equation~\ref{eq:patch-norm}. We use cycle spinning, a stride of 4 pixels and 1000 iterations. Which optimizes "to the end" until there is no relevant change in the posterior log-likelihood.

@@

    We ran the comparison on the whole parameter grid, whoever in the following we can only discuss a subset of the results, which we consider the most relevant. All results are available at \url{https://jolideco.github.io/jolideco-comparison}.

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \script{comparison-scenarios.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of different deconvolution methods for different selected scenarios, as described in Section~\ref{subsec:test-datasets}. The simulation of the data used a fixed background level of, a constant exposure and a Gaussian PSF of \todo{x pixels}, corresponding to the "Chandra" scenario. The left column shows the simulated counts data and the second most column from the left the underlying ground truth to compare. The other column show the reconstructed flux by the different methods and prior assumptions. The methods are described in detail in Section~\ref{subsec:methods}.
                A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of methods on the different source scenarios.
    Figure \ref{fig:comparison-scenarios} shows the comparison of the methods for a selected subset of scenarios A1, B3, C3 and D3.

    On scenario A1 all methods show acceptable to excellent results for the isolated point sources in the upper right of the image. This demonstrates all methods take into account the PSF and background model model correctly. The early stopped uniform prior still shows an extension for those cases, because the optimization process has not finished. The \jolideco with the Zoran-Weiss patch prior also results in less sharp images, because the prior has been learned on natural images of landscapes, objects, every day scenes etc. which do not include point sources. The \jolideco with the prior learned from GLEAM data, which also includes point sources, shows much improved results.

    In the lower left of the image the point sources become fainter and closer, challenging all methods in succeeding to separate those as independent sourcea. The uniform prior optimized "to end" shows excellent results, still separating the sources.

    For scenario B3 all method recover the point sources arpund the central extend source. 
    However large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected the uniform prior decompose the emission into many point sources, neglecting the spatial correlation between them. \textbf{Pylira} shows slightly improved results with the gaps between the bright emission peeks being more filled, however still lacks a convincing result into reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the 
    sources almost perfectly, visually indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a visible extension in the surrounding point sources.

    Scenario C3 challenges the methods to reconstruct a weak extended source under four brighter point sources. While again all methods reconstruct the point sources well, the visual differences are large for the weal underlying extended emission.

@@

    \subsubsection{Comparing Signal-to-Noise Scenarios}
    %
    \begin{figure*}
        \script{comparison-signal-noise.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-signal-noise.pdf}
            \caption{
                Comparison of the \jolideco method with the GLEAM v0.1 prior for different background levels and signal strength represented by scenario D1-5. The images are grouped by background levels with each column group showing the counts data in the left column and the reconstruction in the right column. The rows show the scenarios D1-5 as described in Section~\ref{subsec:test-datasets}. The background levels correspond to \qty[mode = text]{0.001}{cts / pixel}, \qty[mode = text]{0.01}{cts / pixel} and \qty[mode = text]{0.1}{cts / pixel}. The complete results for the simulations and additional information are available at \url{https://jolideco.github.io/jolideco-comparison}. \todo{the background handling does not yet seem right...maybe we should let the background norm free}
            }
            \label{fig:comparison-signal-noise}
        \end{centering}
    \end{figure*}
    %
    \todo{We should give a single number for the S/N, how to compute? Per source, like spiral ring and point sources...}

    \todo{Should we leave the background norm free?}
    Figure \ref{fig:comparison-signal-noise} shows the comparison for the method "Jolideco GLEAM v0.1" for varying signal to noise ratio for the "D" scenario.

@@

    \begin{itemize}
        \item as expected the reconstruction quality improves with stronger signal
        \item there seems to be a "jump" in the reconstruction quality of the point sources. The weakest is much more blurry, but all the others look fine.
        \item the method "fails" at the highest background level.
    \end{itemize}

@@

    \subsection{Combined \xmm and \chandra Observation}
    As a second application example we combine data from \chandra and \xmm.
    \url{https://github.com/jolideco/jolideco-chandra-nustar-examples}

    \begin{figure*}
        \script{example-xmm-chandra.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/example-xmm-chandra.pdf}
            \caption{
                Chandra Example
            }
            \label{fig:example-xmm-chandra}
        \end{centering}
    \end{figure*}

>>>>

    \subsection{Related Work}
    In literature there are many proposed methods to solve the un-blind deconvolution problem for images affected by Poisson noise. All methods rely fundamentally on modeling an image generating process as a linear system. Given a known point spread function a prediction of the discrete measured image can be obtained via \textit{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the likelihood function.

    The standard baseline method for image de-convolution was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \textit{RL}). The method involves estimating the original image by iteratively convolving the observed image with the PSF and dividing by the result of convolving the estimated image with the transposed PSF. The iterations continue until the estimated image converges to a satisfactory solution.

    In practice the RL method was found to have multiple, partly related issues: first there is no objective criterion when to stop the iteration. Usually the iteration was stopped when the reconstructed image seemed plausible to the eye. Second, when the number of iterations was too high, the method shows a tendency to decompose regions of constant brightness into a set of bright emission peaks (\textit{speckles}). This problem was e.g. described and studied by \cite{Reeves1995} and \cite{Fish1995}. Lastly the RL method does not provided an associated estimate of the uncertainties of the reconstruction, which is desirable especailly in the Poisson domain to asses the significance of specific features found in the reconstruction.

    To address the question of when to stop the RL iteration \cite{Reeves1995} proposed to use cross validation. The authors of \citep{Bi1994} proposed a bootstrapped based assessment of the significance of reconstructed image as as a stopping rule for RL.

    To address the issue with RL decomposing extended emission structures into groups of point source, multiple additions to the RL method were proposed. Most importantly this lead to the introduction of regularised RL methods.
    such as use of total variation \citep{Dey2006} or sparseness \todo{add reference}.

    Multi-scale LIRA prior has been used in image deconvolution to incorporate error estimates and Bayesian sampling. This prior has been used in various studies, including those by \cite{Esch2004} and \cite{Connors2011}. However, the prior is ad-hoc and lacks physical information or assumptions, limiting its applicability.

@@

    An alternative approach to solve the problem in the framework of information field theory. Recently, image deconvolution using physical assumptions and decomposition into point and diffuse flux in 3D/4D has gained attention. \cite{Pumpe2018} and \cite{Selig2015} proposed priors based on physical assumptions, which have been shown to work well for large-scale emission. However, the smoothness assumption in these methods can lead to artifacts in the galactic plane.

@@

    \cite{Zoran2011} build on very similar assumption for natural images.
    Instead of assuming a fixed local correlation structures, such as smoothness
    they proposed to learn the local correlation structure on millions of patches
    extracted from natural images.

    \cite{Bouman2016} later adapted the patch prior reconstruction to be used

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the \textit{de facto} standard for de-blurring and super-resolution of natural images \todo{reference review?}. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

@@

    \todo{did you know this sum corresponds to the loop over data sets in machine learning...?}

@@

    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the same patch based image prior approach as introduced by \cite{Zoran2011}. We use a pre-trained Gaussian Mixture Model (GMM) to evaluate and approximate a patch based prior for the reconstructed image $\mathbf{x}$.

    One convenient property of the GMM is the possibility to evaluate its likelihood in closed form. For a given single patch $\mathbf{x}_n$ it is given by:

    \begin{equation}
        \label{eq:gmm}
        P_{GMM}(\mathbf{x}_n) = \sum_{k=1}^K \pi_k \mathcal{N}(\mathbf{x}_n| \boldsymbol{\mu}_k, \boldsymbol{\Sigma}_k^2)
    \end{equation}

    Where $\mathcal{N}$ is a Gaussian distribution, $K$ the total number of components, $\boldsymbol{\mu}_k$ the vector of means for a given component $k$, $\boldsymbol{\Sigma}_k$ the corresponding covariance matrix
    and $\pi_k$ the weight of the component. The different weights are required to sum to unity such that $\sum_{k=1}^N\pi_k = 1$.

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch
    from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the GMM component becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$.

@@

    with radio astronomy data.
    They have shown that the reconstructed image only weakly depends on the choice
    of the reference data on which the the patch prior is learned. They found
    equivalent results for GMMs learned on natural images and specifically
    simulated images of black hole ring structures.

@@

    We also use \texttt{Numpy}~\citep{Numpy2020} for handling data arrays and
    \texttt{Matplotlib}~\citep{Hunter2007} for plotting.
    To learn the Gaussian Mixture Models from patches we use the standard implementation
    provided in the \texttt{Scikit-Learn}~\citep{Skimage2014} package. To handle the 
    serialisation of images to the \textit{FITS} data format and to handle world coordinate
    systems (WCS) we use \texttt{Astropy}~\citep{Astropy2018}.
    For optimization and internal data handling we use \texttt{PyTorch}.

@@

    \vlk{We have devised four distinct arrangements of combinations of point and extended sources of different shapes to test the algorithms.  The arrangements are 
    \begin{itemize}
        \item[(A)] {\bf asterism:} A set of four point sources arranged around an extended Gaussian source
        \item[(B)] {\bf points:} Point sources arranged with different separations
        \item[(C)] {\bf shield:} Disk shaped flat extended source with superposed point sources and linear jet-like features
        \item[(D)] {\bf spiral:} Extended thin double-spiral with a flat disk at the center and point sources adjacent to it
    \end{itemize}
    The arrangements are shown in Figure~TBD, and the details of the locations and brightness of each component is listed in Table~TBD.}

@@

    \subsection{Methods}
    \label{subsec:methods}
    We compare the performance of \jolideco with different prior assumptions and against the \textit{LIRA} method introduced by \cite{Esch2004}. More specifically we use the following methods:

@@

        \item \textbf{Jolideco (Patch, GLEAM v0.1)}: \jolideco with a patch prior learned on GLEAM data, described in Section~\ref{sssec:gleam-radio-data}. It also uses the standardised patch normalisation as described by Equation~\ref{eq:patch-norm}. We use cycle spinning, a stride of 4 pixels and 1000 iterations. Which optimizes "to the end" until there is no relevant change in the posterior log-likelihood. In addition we use and upsampling factor of 2.

@@

    We ran the comparison on the whole parameter grid of methods, source scenarios, background levels and instrument scenarios. However in the following we only discuss a subset of the results, which we consider the most relevant. All results are available at \url{https://jolideco.github.io/jolideco-comparison}.

    \subsection{Results}
    \subsubsection{Comparing Methods}
    %
    \begin{figure*}
        \script{comparison-scenarios.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-scenarios.pdf}
            \caption{
                Comparison of different deconvolution methods for different selected scenarios, as described in Section~\ref{subsec:test-datasets}. The simulation of the data used a fixed background level of, a constant exposure and a Gaussian PSF of \todo{x pixels}, corresponding to the "Chandra" scenario. The left column shows the simulated counts data and the second most column from the left the underlying ground truth to compare. The other column show the reconstructed flux by the different methods and prior assumptions. The methods are described in detail in Section~\ref{subsec:methods}.
                A more detailed representation of the results is available at \url{https://jolideco.github.io/jolideco-comparison}.
            }
            \label{fig:comparison-scenarios}
        \end{centering}
    \end{figure*}
    %
    First we compare the performance of methods on the different source scenarios.
    Figure \ref{fig:comparison-scenarios} shows the comparison of the methods for a selected subset of scenarios A1, B3, C3 and D4.

    On scenario A1 all methods show acceptable to excellent results for the isolated point sources in the upper right of the image. This demonstrates all methods take into account the PSF and background model model correctly. The early stopped uniform prior still shows an extension for those cases, because the optimization process has not finished. Optimizing for 1000 iterations clearly improves the result and shows almost perfect point sources corresponding to the size of approximately one pixel. The \jolideco method with the Zoran-Weiss patch prior results in less sharp images, because the prior has been learned on natural images of landscapes and every day objects and scenes, which do not include point sources. The \jolideco method with the prior learned from GLEAM data, which also includes point sources, shows much improved results. However not as sharp as the results obtained with the uniform prior.

    In the lower left of the image the point sources become fainter and move closer, challenging all methods in succeeding to separate those as independent sources. The uniform prior optimized "to end" shows excellent results, still separating the sources.

    For scenario B3 all method recover the point sources and the central extend source. 
    However large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected the uniform prior decompose the emission into many point sources, neglecting the spatial correlation between them. \textbf{Pylira} shows slightly improved results with the gaps between the bright emission peeks being more filled, however still lacks a convincing result into reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the 
    sources almost perfectly, visually indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a visible extension in the surrounding point sources.

    Scenario C3 challenges the methods to reconstruct a weak extended source under four brighter point sources. While again all methods reconstruct the point sources well, the visual differences are large for the weak underlying extended emission. The uniform prior with 1000 iterations

@@

    \subsubsection{Comparing Signal-to-Noise Scenarios}
    %
    \begin{figure*}
        \script{comparison-signal-noise.py}
        \begin{centering}
            \includegraphics[width=\linewidth]{figures/comparison-signal-noise.pdf}
            \caption{
                Comparison of the \jolideco method with the GLEAM v0.1 prior for different background levels and signal strength represented by scenario D1-5. The images are grouped by background levels with each column group showing the counts data in the left column and the reconstruction in the right column. The rows show the scenarios D1-5 as described in Section~\ref{subsec:test-datasets}. The background levels correspond to \qty[mode = text]{0.001}{cts / pixel}, \qty[mode = text]{0.01}{cts / pixel} and \qty[mode = text]{0.1}{cts / pixel}. The complete results for the simulations and additional information are available at \url{https://jolideco.github.io/jolideco-comparison}. \todo{the background handling does not yet seem right...maybe we should let the background norm free}
            }
            \label{fig:comparison-signal-noise}
        \end{centering}
    \end{figure*}
    %
    As a second experiment we explored the dependency of the results on background level and signal strength. Figure~\ref{fig:comparison-signal-noise} shows the results for the \jolideco method with the GLEAM prior in the scenario D1-5. As expected the reconstruction quality improves with signal strength. The method seems to converge to the ground truth for very high signal-to-noise ratio.

    \todo{We should give a single number for the S/N, how to compute? Per source, like spiral ring and point sources...}

@@

    \begin{itemize}
        \item 
        \item there seems to be a "jump" in the reconstruction quality of the point sources. The weakest is much more blurry, but all the others look fine.
        \item the method "fails" at the highest background level.
    \end{itemize}

@@

    % \subsection{Combined \xmm and \chandra Observation}
    % As a second application example we combine data from \chandra and \xmm.
    % \url{https://github.com/jolideco/jolideco-chandra-nustar-examples}

    % \begin{figure*}
    %     \script{example-xmm-chandra.py}
    %     \begin{centering}
    %         \includegraphics[width=\linewidth]{figures/example-xmm-chandra.pdf}
    %         \caption{
    %             Chandra Example
    %         }
    %         \label{fig:example-xmm-chandra}
    %     \end{centering}
    % \end{figure*}
